% FEISTEL white paper

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[twoside,twocolumn]{article}

\usepackage{blindtext} % Package to generate dummy text throughout this template 

\usepackage[sc]{mathpazo} % Use the Palatino font
\usepackage[T1]{fontenc} % Use 8-bit encoding that has 256 glyphs
%\linespread{1.05} % Line spacing - Palatino needs more space between lines
\usepackage{microtype} % Slightly tweak font spacing for aesthetics
\usepackage{eufrak}
\usepackage{graphicx} % For \scalebox

\usepackage[english]{babel} % Language hyphenation and typographical rules

\usepackage[hmarginratio=1:1,top=32mm,columnsep=20pt]{geometry} % Document margins
\usepackage[hang, small,labelfont=bf,up,textfont=it,up]{caption} % Custom captions under/above floats in tables or figures
\usepackage{booktabs} % Horizontal rules in tables

\usepackage{lettrine} % The lettrine is the first enlarged letter at the beginning of the text

\usepackage{enumitem} % Customized lists
\setlist[itemize]{noitemsep} % Make itemize lists more compact

\usepackage{abstract} % Allows abstract customization
\renewcommand{\abstractnamefont}{\normalfont\bfseries} % Set the "Abstract" text to bold
\renewcommand{\abstracttextfont}{\normalfont\small\itshape} % Set the abstract itself to small italic text

\usepackage{titlesec} % Allows customization of titles
\renewcommand\thesection{\Roman{section}} % Roman numerals for the sections
\renewcommand\thesubsection{\arabic{subsection}} % roman numerals only for subsections
\titleformat{\section}[block]{\Large\scshape\centering}{\thesection.}{1em}{} % Change the look of the section titles
\titleformat{\subsection}[block]{\large\scshape}{\thesubsection.}{1em}{} % Change the look of the section titles

\usepackage{fancyhdr} % Headers and footers
\pagestyle{fancy} % All pages have headers and footers
\fancyhead{} % Blank out the default header
\fancyfoot{} % Blank out the default footer
\fancyhead[C]{Feistel Cipher with Hash Round Function $\bullet$ Cyril Dever} % Custom header text
\fancyfoot[RO,LE]{\thepage} % Custom footer text
\setlength{\headheight}{14pt}

\usepackage{titling} % Customizing the title section

\usepackage{hyperref} % For hyperlinks in the PDF

\usepackage[symbol]{footmisc} % To use special character in footnote
\renewcommand{\thefootnote}{\arabic{footnote}}

\usepackage{outlines}
\usepackage[font=itshape]{quoting}

\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\SetKw{Continue}{continue}
\SetKw{KwBy}{by}

%----------------------------------------------------------------------------------------
%	FUNCTIONS
%----------------------------------------------------------------------------------------

\newcommand{\ceil}[1]{\left\lceil #1 \right\rceil}
\newcommand{\floor}[1]{\left\lfloor #1 \right\rfloor}
\newcommand{\bsfnote}{\textsuperscript{*}} % for reference to the base64 string note
\newcommand{\hexnote}{\textsuperscript{$\dagger$}} % for reference to the hex string note
\newcommand{\mod}[1]{\ \mathrm{mod}\ #1}

%----------------------------------------------------------------------------------------
%	LISTINGS
%----------------------------------------------------------------------------------------

\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}

\theoremstyle{remark}
\newtheorem*{remark}{Note}
\newtheorem*{recall}{Recall}

%----------------------------------------------------------------------------------------
%	FIGURES
%----------------------------------------------------------------------------------------

\usepackage{tikz}
\usepackage{caption}

\usetikzlibrary{shapes.geometric, arrows, calc, positioning}

\tikzstyle{startstop} = [rectangle, rounded corners, minimum width=3cm, minimum height=1cm,text centered, draw=black]
\tikzstyle{io} = [trapezium, trapezium left angle=70, trapezium right angle=110, minimum width=3cm, minimum height=1cm, text centered, text width=1.7cm, inner sep=0.4cm, draw=black]
\tikzstyle{process} = [rectangle, minimum width=3cm, minimum height=1cm, text centered, draw=black]
\tikzstyle{decision} = [diamond, minimum width=3cm, minimum height=1cm, text centered, inner sep=-0.1cm, draw=black]
\tikzstyle{arrow} = [thick,->,>=stealth]
\tikzset{XOR/.style={draw,circle,append after command={
        [shorten >=\pgflinewidth, shorten <=\pgflinewidth,]
        (\tikzlastnode.north) edge (\tikzlastnode.south)
        (\tikzlastnode.east) edge (\tikzlastnode.west)
        }
    }
}

%----------------------------------------------------------------------------------------
%	TITLE SECTION
%----------------------------------------------------------------------------------------

\usepackage[english]{datetime2}
\DTMsavedate{thedate}{2019-11-21} % V0 brings the need for Feistel cipher

\setlength{\droptitle}{-5\baselineskip} % Move the title up

\pretitle{\begin{center}\Large\bfseries}
\posttitle{\end{center}}
\title{Feistel Cipher with Hash Round Function} % Title
\author{%
    \textsc{Cyril Dever}\\ % Name
    \normalsize Edgewhere \\ % Institution
}
% \date{\today} % Leave empty to omit a date
\date{\DTMusedate{thedate}}
\renewcommand{\maketitlehookd}{%
    \begin{abstract}
        \noindent We define an obfuscation tool to secure data with an almost Format-Preserving Encryption process. By implementing a Feistel block 
        cipher with a round function using any robust hashing function, it provides you with a one-way tool to both encrypt and decrypt the data.
    \end{abstract}
}

%----------------------------------------------------------------------------------------

\begin{document}

% Print the title
\maketitle

%----------------------------------------------------------------------------------------
%	ARTICLE CONTENTS
%----------------------------------------------------------------------------------------

\section{Introduction}

\lettrine[nindent=0em,lines=3]{P}rovided you need a robust obfuscation function for protecting your data more than an actual encryption cipher, meet 
our own implementation of the well-known Feistel network algorithm.

It's secure, yet very fast, and it comes with two handy features:
\begin{itemize}
    \item Encryption and decryption both uses the same way to work, ie. the use of only one function is needed for both obfuscating and recovering data;
    \item The end result respects Format-Preserving Encryption (FPE), ie. the length of the output is the same as the one of the output. % // TODO
\end{itemize}

\section{The Algorithm}

\subsection{Formal Description}

We herein define $\mathfrak{F}$ our own implementation of a Feistel block cipher\cite{feistel:hf}. 

We use a balanced implementation, cutting the input data into two equal parts, processing them through our round function (see section 
\ref{roundFunction}), to finally concatenating the end results to form the final obfuscated ciphertext.

It is an almost Format-Preserving Encryption scheme; "almost" because it depends on the size of the input. If the latter is of even length, then the 
output will preserve its size; otherwise, we'd pad it at the start of the process (see section \ref{padding}), making it longer by one character:
\begin{equation}
    \label{fpe}
    \begin{array}{l}
        y \gets \mathfrak{F}(x) \\
        \Rightarrow |y| = \left\{
            \begin{array}{l}
                \textbf{if } x \mod 2 = 0 \textbf{ then } |x| \\
                \textbf{else } |x| + 1 \\
            \end{array}
            \right.
    \end{array}
\end{equation}

Let us start with what we use as the basis for our own implementation: the formal description provided by Wikipedia
\footnote{\url{https://en.wikipedia.org/wiki/Feistel_cipher}} for a Feistel block cipher is as described in Algorithm \ref{algo:feistel}.

Let $N = n+1$ be the number of rounds, $K_{0},K_{1},...,K_{n}$ the keys associated with each round and $F: \omega \times \mathcal{K} \mapsto \omega$ a 
function of the $(words \times keys)$ space to the $words$ space.
\begin{algorithm}
    \KwIn{a message $m$}
    \KwOut{the ciphertext $c$}
    let the encrypted word in step $i$ be $m_i := L_i \mathbin\Vert R_i$ with $m_0 := L_0 \mathbin\Vert R_0$ as the unciphered message; \\
    \For{$i \gets 0$ \KwTo $n$ \KwBy $1$}{
        \qquad $L_{i+1} \gets R_i$; \\
        \qquad $R_{i+1} \gets L_i \oplus F(L_i,K_i)$; \\
    }
    $m_N := L_{n+1} \mathbin\Vert R_{n+1}$; \\
    \Return{$m_N$}
    \caption{Standard Feistel cipher}
    \label{algo:feistel}
\end{algorithm}

\subsection{Padding}
\label{padding}

We could have turned our cipher into a fully FPE-compliant system by forcing the input data to be of even length.

Instead, we kept a smoother approach by deciding we'd add the padding ourselves. That way, our users don't have to worry about this step, only that 
the output might be one character longer than the input (as seen above).

But of course, should you provide data of even length (using your own padding system), then our cipher definitely follows a FPE scheme.

Algorithm \ref{algo:padding} defines our left padding function $P()$.
Let \texttt{PAD\_CHAR} be a padding character\footnote{In our own implementation, we use the UTF-8 \texttt{U+0002} (start-of-text) character.}.
\begin{algorithm}
    \KwIn{a message $m$, \texttt{PAD\_CHAR}}
    \KwOut{the balanced message}
    \eIf{$|m| \mod 2 = 0$}{
        \Return{$m$}
    }{
        \Return{\emph{\texttt{PAD\_CHAR}} $\mathbin\Vert m$}
    }
    \caption{Padding $P$}
    \label{algo:padding}
\end{algorithm}

Algorithm \ref{algo:unpadding} shows its inverse, ie. the unpadding function.

\begin{algorithm}
    \KwIn{a padded message $m$, \texttt{PAD\_CHAR}}
    \KwOut{the unpadded message}
    \eIf{$m[0] =$ \emph{\texttt{PAD\_CHAR}}}{
        \Return{$\mathbin\Vert_{i=1}^{|m|-1} m[i]$}
    }{
        \Return{$m$}
    }
    \caption{Unpadding $P^{-1}$}
    \label{algo:unpadding}
\end{algorithm}

\subsection{Hash Round Function}
\label{roundFunction}

Figure \ref{fig:feistel} provides a graphical representation of our cipher $\mathfrak{F}$ in its entirety.

Our implementation takes its robustness by actually not using one different key per round, but rather by using a well-tested hash function\footnote{
    In our first implementation, we use the \texttt{SHA-256} hash algorithm as $\mathfrak{h}()$ because it is both widely adopted (in particular 
    natively in most browsers) and yet still very secure at the time of this writing.
} $\mathfrak{h}()$ and a single key $K$ in its round function $F$.

\begin{figure}
    \centering\noindent
    \begin{tikzpicture} %see 'https://www.iacr.org/authors/tikz/'
        \centering \noindent
        % First two rounds
        \node[draw,thick,minimum width=1cm] (f1) at ($1*(0,-1.5cm)$)  {$F_1$};
        \node (xor1) [XOR, left of = f1, node distance = 2cm] {};
        \draw[thick,-latex] (f1) -- (xor1);
    
        \node[draw,thick,minimum width=1cm] (f2) at ($2*(0,-1.5cm)$)  {$F_2$};
        \node (xor2) [XOR, left of = f2, node distance = 2cm] {};
        \draw[thick,-latex] (f2) -- (xor2);
        
        \draw[thick,latex-latex] (f1.east) -| +(1.5cm,-0.5cm) -- ($(xor1) - (0,1cm)$) -- ($(xor1.north) - (0,1.5cm)$);
        \draw[thick] (xor1.south) -- ($(xor1)+(0,-0.5cm)$) -- ($(f1.east) + (1.5cm,-1cm)$) -- +(0,-0.5cm);
        
        \draw[thick,latex-] (f2.east) -| +(1.5cm,-0.5cm) -- ($(xor2) - (0,1cm)$);
        \draw[thick] (xor2.south) -- ($(xor2)+(0,-0.5cm)$) -- ($(f2.east) + (1.5cm,-1cm)$);
        
        \draw[thick, densely dotted] ($(f2.east) + (1.5cm,-1cm)$) -- +(0,-0.5cm);
        \draw[thick, densely dotted] ($(xor2) - (0,1cm)$) -- ($(xor2.north) - (0,1.5cm)$);
        
        % Middle text
        \node at (0,-4.5cm) {\scriptsize{for $n$ rounds}};
    
        % Last two rounds
        \node[draw,thick,minimum width=1cm] (f3) at ($3*(0,-1.5cm) + (0, -.75cm)$)  {$F_{n-1}$};
        \node (xor3) [XOR, left of = f3, node distance = 2cm] {};
        \draw[thick,-latex] (f3) -- (xor3);
    
        \node[draw,thick,minimum width=1cm] (f4) at ($4*(0,-1.5cm) + (0, -.75cm)$)  {$F_{n}$};
        \node (xor4) [XOR, left of = f4, node distance = 2cm] {};
        \draw[thick,-latex] (f4) -- (xor4);
        
        \draw[thick,latex-latex] (f3.east) -| +(1.5cm,-0.5cm) -- ($(xor3) - (0,1cm)$) -- ($(xor3.north) - (0,1.5cm)$);
        \draw[thick] (xor3.south) -- ($(xor3)+(0,-0.5cm)$) -- ($(f3.east) + (1.5cm,-1cm)$) -- +(0,-0.5cm);
        
        \draw[thick, densely dotted] ($(f3.east) + (1.5cm,0cm)$) -- +(0cm,0.5cm);
        \draw[thick, densely dotted] (xor3.north) -- +(0cm,0.35cm);
    
        %% Inputs    
        \node (p0) [draw,thick,above of = f1, minimum width=5cm,minimum height=0.5cm,node distance=1cm] {$in$}; 
        \node (l0) [above of = xor1,node distance=2cm] {$L$};
        \node (r0) [right of = l0, node distance = 4cm] {$R$};
        \draw[thick,-latex] (l0 |- p0.south) -- (xor1.north);
        \draw[thick] ($(f1.east)+(1.5cm,0)$) -- +(0,0.75cm);
    
        \draw[thick,latex-] (l0 |- p0.north) -- (l0);
        \draw[thick,latex-] (r0 |- p0.north) -- (r0);
    
        %% Outputs
        \node (p4) [draw,thick,below of = f4, minimum width=5cm,minimum height=0.5cm,node distance=1cm] {$out$}; 
        \draw[thick,latex-latex] (f4.east) -| +(1.5cm,-0.75cm);
        \draw[thick,-latex] (xor4.south) -- ($(xor4)+(0,-0.75cm)$);    
    \end{tikzpicture}
    \caption{Feistel block cipher $\mathfrak{F}$}
    \label{fig:feistel}
\end{figure}

The round function $F$ thus consists in taking the right side $R$ at each round and apply to it two operations:
\begin{itemize}
    \item Shift $K$ by the number of round;
    \item Add $R$ to the masked key $K'$ (of the shifted $K$);
    \item Hash the result $R'$ through $\mathfrak{h}()$.
\end{itemize}

\subsubsection{Shifting the key}

To shift the passed key $K$ by one character at each round, we use the shifting function $S()$.

Let $substr(x, start)$ be a function that keeps the substring of the passed $x$ from $start$ to end.
\begin{equation}
    \label{eq:shifting}
    \begin{array}{rl}
        S: \mathcal{K} \times \mathbb{N} &\to \mathcal{K} \\
            (K, i) &\mapsto substr(K \ll i, 1) \mathbin\Vert K[0] \\
    \end{array}
\end{equation}

That way, we build a "new" key from $K$ for $|K|$ rounds, adding security to our round function.

\subsubsection{Masking the new key}

To enable the XOR part of the Feistel cipher, we have to apply a "masking" operation $\mu()$ on the input key $K$ to make it of length $l = |R|$:
\begin{small}
    \begin{equation}
        \label{eq:masking}
        \begin{array}{rl}
            \mu: \mathcal{K} \times \mathbb{N} &\to \mathcal{K} \\
            (K, l) &\mapsto K' := \left\{
                \begin{array}{l}
                    \textrm{if } |K| \geq l, \sum_{i=1}^l K[i] \\ \\
                    \sum_{i=1}^l (K \times \ceil{|K| \div l})[i] \\
                \end{array}
            \right.
        \end{array}
    \end{equation}
\end{small}
If the key $K$ is too long, the masking function $\mu()$ eventually cuts it, ie. only keeping the $l$-th first bytes.
And if it is too short, it multiplies it the needed number of times and cut the concatenation to the desired length $l$.

\subsubsection{Adding parts}

At each round, we add the masked key $K'$ with the right part of the previous round $R$ through the function $A()$ described in Algorithm 
\ref{algo:addition}.

Let $charcode$ be the UTF-8 character code of the concerned byte.
\begin{algorithm}
    \KwIn{$R$, $K' \gets \mu(K, |R|)$}
    \KwOut{$R'$}
    initialize $R' \gets \emptyset$ of length $|R|$; \\
    \ForEach{charcode $i \in R$ and $i \in K'$}{
        $R'[i] := R[i] + K'[i]$;
    }
    \Return{$R'$}
    \caption{Addition function $A$}
    \label{algo:addition}
\end{algorithm}

For example, the addition of \texttt{a} with \texttt{b} gives:
\texttt{a} $\gets 61$,  \texttt{b} $\gets 62 \Rightarrow \texttt{a} + \texttt{b} \gets 123 \mapsto \texttt{b01111011}$.

\subsubsection{Wrapping it all up}

We define the final round function $F$ at round $i$ as the hash of the previous addition, the result we XOR with the left part $L$ of the previous 
round to form the new basis for the next round where $L$ and the output of $F$ are switched.
\begin{small}
    \begin{equation}
        \label{eq:roundFunction}
        \begin{array}{rl}
            F: \omega \times \mathcal{K} \times \mathbb{N}  &\to \omega \\
                (R, K, i)  &\mapsto \mathfrak{h} \left[ A \Bigg(R, \mu \Big( S(K,i), |R| \Big) \Bigg) \right] \\
        \end{array}
    \end{equation}
\end{small}

$F$ is applied at every round. And our implementation eventually unpads the result of $\mathfrak{F}$ by adding a final $P^{-1}()$ step at the end.

\subsection{Full cipher}

The last parameter of the whole cipher $\mathfrak{F}$ is the number of rounds $N$.
Note that it has been proved \cite{Permutations:lr} that, for such an implementation of the Feistel block cipher, four rounds of permutations are 
enough to make it "strong"\footnote{but we usually use at least 10 rounds}.

We finally define the full cipher $\mathfrak{F}$ that respects Figure \ref{fig:feistel} with $F_i = F(R, K, i)$ at round $i$ as follows:
\begin{equation}
    \label{eq:fullCipher}
    \begin{array}{rl}
        \mathfrak{F}: \omega \times \mathcal{K} \times \mathbb{N} &\to \omega \\
                        (m, K, N) &\mapsto \mathfrak{F}(P(m), K, N) \\
    \end{array}
\end{equation}

\begin{recall}
    One of the main advantage of using this Feistel block cipher construction is that encryption and decryption are similar:$$
        out = \mathfrak{F}(in, K_\gamma, n) \iff in = \mathfrak{F}(out, K_\gamma, n)
    $$
\end{recall}

\section{Implementation}

We created two different implementations for now: one in JavaScript\footnote{\url{https://npmjs.org/package/feistel-ceipher}} and one in Go
\footnote{\url{https://github.com/cyrildever/feistel}}.

On both environments, our latest tests show no significant impact with a $10$ round cipher (a few dozens of nanoseconds at most). The results are in 
fact mostly impacted by the speed of the used hash function on the machine it is run (and obviously a little slower in the browser of an ordinary 
PC).

\vfill\eject % To force break column if need be
\tableofcontents % Uncomment to add a table of contents

%----------------------------------------------------------------------------------------
%	REFERENCE LIST
%----------------------------------------------------------------------------------------

\begin{thebibliography}{99} % Bibliography

\bibitem[1]{feistel:hf}
Horst Feistel. \emph{Cryptography and Computer Privacy}, Scientific American, 1973.

\bibitem[2]{Permutations:lr}
Michael Luby, Charles Rackoff. \emph{How to Construct Pseudorandom Permutations from Pseudorandom Functions}, SIAM Journal on Computing, 1988.

\end{thebibliography}

%----------------------------------------------------------------------------------------

\end{document}